Human Factors Influencing Visual Statistical Inference

### 2015-08-28 sociology_paper.Rnw
references
references added

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms showing simulated poll results for the Democratic presidential candidate using  polling averages collected from two different sources, Freedom's Lighthouse Averages (top), RealClearPolitics (bottom).  If the polls matched the electoral result the simulated results would be centered at the electoral result. This happens for the RealClearPolitics polls (bottom) but not for Freedom's Lighthouse.  For the latter, only 5.33\% of simulated results were as high as the actual election result, which converts to a $p$-value of 0.0533, indicating an almost statistically significant difference. It hints at the polls from this source under-estimating the vote for Obama. This source was used to generate the null plots in the lineup in Figure~\ref{fig:elect}. The $p$-value from the lineup protocol, was 0.0077, which more strongly concludes that these polls mismatched the actual result.}
\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.%the election example, above, the scenario was explained after the lineup question, in the text.

@@

In experiments~4,~5 and~6, different visual test statistics were used on the same data, providing the ability to examine the effect of these choices. Results are published in \cite{heike:2012} and \cite{boxplots:arxiv}.

@@

Model~\eqref{manova} was fit to the data to test, if  detection rates were significantly different for the different locations. For this we use the {\it manova()} function provided in the {\it stats} package of the software R \citep{R:2012}. The results are shown in Table~\ref{tbl:manova}. The $p$-values for both the Interaction and the Genotype effect are much bigger than the conventional threshold of 0.05 suggesting that location of the actual data plot in the lineup does not significantly affect performance.

>>>>

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms showing simulated poll results for the Democratic presidential candidate using  polling averages collected from two different sources, Freedom's Lighthouse Averages (top), RealClearPolitics (bottom).  If the polls matched the electoral result the simulated results would be centered at the electoral result. This happens for the RealClearPolitics polls (bottom) but not for Freedom's Lighthouse.  For the latter, only 5.33\% of simulated results were as high as the actual election result, which converts to a $p$-value of 0.0533, indicating an almost statistically significant difference. It hints at the polls from this source under-estimating the vote for Obama. This source was used to generate the null plots in the lineup in Figure~\ref{fig:elect}. The $p$-value from the lineup protocol, was 0.0077, which more strongly concludes that these polls mismatched the actual result.}
\end{figure}
The lineup of Figure~\ref{fig:elect}, as well as the other lineups in the manuscript are created using the nullabor package by \citep{nullabor}.

@@

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis \citep{vinference}.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.%the election example, above, the scenario was explained after the lineup question, in the text.

@@

In experiments~4,~5 and~6, different visual test statistics were used on the same data, providing the ability to examine the effect of these choices, the use of lineup tests for comparisons and two examples are discussed in \citet{heike:2012}

@@

Model~\eqref{manova} was fit to the data to test, if  detection rates were significantly different for the different locations. For this we use the {\it manova()} function provided in the {\it stats} package of the software R \citep{R}. The results are shown in Table~\ref{tbl:manova}. The $p$-values for both the Interaction and the Genotype effect are much bigger than the conventional threshold of 0.05 suggesting that location of the actual data plot in the lineup does not significantly affect performance.
